---
title: "voluModel: Modeling Species Distributions in Three Dimensions"
author:
- Hannah L. Owens
- Carsten Rahbek

date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/hannahlowens/voluModel/workflows/R-CMD-check/badge.svg)](https://github.com/hannahlowens/voluModel/actions)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Codecov test coverage](https://codecov.io/gh/hannahlowens/voluModel/branch/main/graph/badge.svg)](https://codecov.io/gh/hannahlowens/voluModel?branch=main)
<!-- badges: end -->

## Summary

This package is designed to model species distributions in three dimensions. This is useful in an aquatic context when the organisms one wishes to model can be found across a wide range of depths in the water column.

Please cite `voluModel`. Run the following to get the appropriate citation for the version you're using:

```{r}
citation(package = "voluModel")
```

## Installation

Currently, there is only a development version.

```{r eval=FALSE}
devtools::install_github("hannahlowens/voluModel")
```

```{r}
library("voluModel")
```

## Getting Started

* Vignettes coming soon.

## Meta

* Please [report any issues or bugs](https://github.com/hannahlowens/voluModel/issues).
* License: GPL-3
* Get citation information for `voluModel` in R using `citation(package = 'voluModel')`
